On the validity of the Born approximation for beyond-Gaussian weak lensing observables

### 2017-01-17 Paper/paper.tex
removed complex conjugate
add NYU affiliation
edit paper after referee comments

\author{Zolt\'an Haiman}
\affiliation{Department of Astronomy, Columbia University, New York, NY 10027, USA}

@@

\section{Introduction}
%
Weak gravitational lensing (WL) is a promising observational technique to probe the standard $w$CDM model of the universe, and can help in constraining the Dark Energy equation of state parameters \citep{wlreview}. Accurate forward modeling of WL observables is crucial for parameter estimation purposes. This however is a challenging problem due to the non--linear nature of WL image fields. Numerical efforts for feature forward modeling are based on ray--tracing simulations that make use of the multi--lens--plane algorithm \citep{RayTracingJain,RayTracingHartlap} to predict cosmic shear from outputs of $N$--body simulations. Approximate techniques based on the Born approximation allow for a computationally faster, but potentially inaccurate, forward modeling \citep{RayTracingHartlap} which requires only knowledge of the matter density contrast integrated along the observer unperturbed line of sight. A variety of groups have studied the validity of the Born approximation for forward modeling of convergence power spectra \citep{HirataKrause} and bi--spectra \citep{WLBispectrumDodelson}, spectra of WL cosmic flexions \citep{BornFlexion} and, in more recent work, CMB lensing bi--spectra \citep{CMBPrattenLewis}. In this work, we study the validity of the Born approximation for forward modeling a subset of low--order moments of the real--space convergence field.

@@

Our image feature choice includes the $\kappa$ power spectrum $P^{\kappa\kappa}(\ell)$, defined as 
\begin{equation}
\label{par:powerdef}
\langle\tilde{\kappa}(\pmb{\ell})\tilde{\kappa}^*(\pmb{\ell}')\rangle = (2\pi)^2\delta^D(\pmb{\ell}+\pmb{\ell}')P^{\kappa\kappa}(\ell),
\end{equation}
%
where the expectation value is taken over $\pmb{\ell}$ modes with the same magnitude $\ell=\vert\pmb{\ell}\vert$ and $\delta^D$ is the Dirac delta function. Because the $\kappa$ field is non--Gaussian, higher order statistics such as higher real space $\kappa$ moments have been shown to contain complementary information in addition to the one already supplied by the power spectrum \citep{MinkPetri,CFHTMink,NG-Jain1,NG-Jain2}. We consider the following sets of $\kappa$ moments $\pmb{\mu}_i$:

@@

In this subsection, we compare the accuracy of the Born approximation in predicting WL image features, in particular we focus on the $\kappa$ power spectrum and the 9 moments described in equation (\ref{par:moments}). Figure \ref{fig:psResiduals} shows the auto power spectra of $\kappa_{\rm born},\kappa_{\rm ll},\kappa_{\rm geo}$ as defined in equation (\ref{sim:pb2}). At lowest order the lensing potential contributes to the $\kappa$ power spectrum quadratically. The first post--Born correction to $P^{\kappa\kappa}$ comes in at $O(\Phi^3)$ and is given by
\begin{equation}
\label{res:powercorr}
P^{\kappa\kappa} = P^{\kappa\kappa}_{\rm born,born} + 2P^{\kappa\kappa}_{\rm born,ll} + 2P^{\kappa\kappa}_{\rm born,geo} + O(\Phi^4),
\end{equation}
%
where we defined the $\kappa$ cross spectra as
\begin{equation}
\label{res:powercross}
\langle\tilde{\kappa}_a(\pmb{\ell})\tilde{\kappa}_b^*(\pmb{\ell}')\rangle = (2\pi)^2\delta^D(\pmb{\ell}+\pmb{\ell}')P_{a,b}^{\kappa\kappa}(\ell).
\end{equation} 
%
Figure \ref{fig:psResiduals} shows the magnitudes of the $O(\Phi^3)$ post--Born corrections to the convergence power spectrum. A similar analysis can be carried out for the various skewness and kurtosis moments of the $\kappa$ field. The main contributions to the skewness and kurtosis enter at $O(\Phi^3)$ and $O(\Phi^4)$, respectively. The first post--Born corrections, on the other hand, enter at $O(\Phi^5),O(\Phi^6)$. In more detail, we have

@@

\section{Discussion}
\label{sec:discuss}
% 
In this section, we discuss our main findings. Figure \ref{fig:csample} shows that the dominant post--Born correction to the $\kappa$ reconstruction comes from the geodesic term (eq.~\ref{sim:gp}). This can be seen by looking at the lens-lens term, which is over an order of magnitude smaller than the geodesic term, and by looking at the full $\kappa-\kappa_{\rm born}$ residuals, which show a very similar overall level, as well as structural detail, to the ones found in the $\kappa_{\rm geo}$ map. The reader might have noticed the curious alignment feature in Figure \ref{fig:csample}, in which all the dipolar structures in $\kappa_{\rm geo}$ appear to be horizontally aligned throughout the field of view. We believe that this alignment is not a numerical effect, but is originated physically by the $\nabla\delta$ terms in eq.~(\ref{sim:gp}). Early in the line--of--sight integration, the field of view covers a very small section of the simulation box ($\sim7\,{\rm Mpc}$ for the first lens), within which dipolar structures are aligned. These alignments survive after the full integration is completed due to the $\nabla\delta\cdot\nabla\Phi$ term, which couples lenses at different redshifts. To provide evidence that the closest lenses to the observer are responsible for the alignment features, we performed the line--of--sight integration for $\kappa_{\rm geo}$ by rotating the first 5 lenses by 90 degrees. This is sufficient to coherently flip the dipolar structures from horizontal to vertical throughout the field of view.

@@

\begin{itemize}
\item The lowest--order post--Born corrections to the convergence power spectrum, skewness and kurtosis are sufficient to fully account for the discrepancies between ray--tracing and the Born approximation.
\item Using the \ttt{LensTools} software package, Born integration costs 4 times less than ray--tracing, and consumes about half as much memory.
\item The Born approximation for the $\kappa$ power spectrum leads to negligible parameter bias for an LSST--like survey, even in the noiseless case.
\item Fitting an observation with Born--approximated higher $\kappa$ moments leads to significant bias in the $(\Omega_m,w_0,\sigma_8)$ triplet in the noiseless case. However, this bias becomes negligible when galaxy shape noise is added, due to an increased scatter in parameter estimates. This conclusion holds even for future space--based surveys that can probe galaxy angular densities as high as $60\,{\rm arcmin}^{-2}$.    
\end{itemize}
%
In this work we examined the validity of the Born approximation for WL galaxy surveys, but a similar study could in principle be carried for the lensing potential reconstruction from CMB temperature and polarization data. Lensing of the CMB probes structures over a wider range of redshifts and hence CMB lensing observables can be expected to be more Gaussian distributed than their galaxy lensing counterparts. This suggests the possibility that non--Gaussian features in CMB lensing data could come from post--Born corrections of $O(\Phi^2)$ rather than inherent non--Gaussianity in the Born $O(\Phi)$ terms. This possibility has been suggested by \citep{CMBPrattenLewis}, who looked at the CMB lensing $\kappa$ bi--spectrum. Because of the high significance with which non--Gaussianities in the CMB lensing potential can be detected with future Stage IV experiments \citep{CMBLNG}, we propose to investigate post--Born corrections to CMB lensing observables in future work.

>>>>

\author{Zolt\'an Haiman}
\affiliation{Department of Astronomy, Columbia University, New York, NY 10027, USA}
\affiliation{Department of Physics, New York University, New York, NY 10003, USA}

@@

\section{Introduction}
%
Weak gravitational lensing (WL) is a promising observational technique to probe the standard $w$CDM model of the universe, and can help in constraining the Dark Energy equation of state parameters \citep{wlreview}. Accurate forward modeling of WL observables is crucial for parameter estimation purposes. This however is a challenging problem due to the non--linear nature of WL image fields. Numerical efforts for feature forward modeling are based on ray--tracing simulations \textbf{\color{red} that make use of the multi--lens--plane algorithm \citep{RayTracingJain,RayTracingHartlap,RayTracingPN}, or the direct approach \citep{RayTracingBarreira},} to predict cosmic shear from outputs of $N$--body simulations. Approximate techniques based on the Born approximation allow for a computationally faster, but potentially inaccurate, forward modeling \citep{RayTracingHartlap} which requires only knowledge of the matter density contrast integrated along the observer unperturbed line of sight. A variety of groups have studied the validity of the Born approximation for forward modeling of convergence power spectra \citep{HirataKrause} and bi--spectra \citep{WLBispectrumDodelson}, spectra of WL cosmic flexions \citep{BornFlexion} and, in more recent work, CMB lensing bi--spectra \citep{CMBPrattenLewis}. In this work, we study the validity of the Born approximation for forward modeling a subset of low--order moments of the real--space convergence field.

@@

Our image feature choice includes the $\kappa$ power spectrum $P^{\kappa\kappa}(\ell)$, defined as 
\begin{equation}
\label{par:powerdef}
\langle\tilde{\kappa}(\pmb{\ell})\tilde{\kappa}(\pmb{\ell}')\rangle = (2\pi)^2\delta^D(\pmb{\ell}+\pmb{\ell}')P^{\kappa\kappa}(\ell),
\end{equation}
%
where the expectation value is taken over $\pmb{\ell}$ modes with the same magnitude $\ell=\vert\pmb{\ell}\vert$ and $\delta^D$ is the Dirac delta function. Because the $\kappa$ field is non--Gaussian, higher order statistics such as higher real space $\kappa$ moments have been shown to contain complementary information in addition to the one already supplied by the power spectrum \citep{MinkPetri,CFHTMink,NG-Jain1,NG-Jain2}. We consider the following sets of $\kappa$ moments $\pmb{\mu}_i$:

@@

In this subsection, we compare the accuracy of the Born approximation in predicting WL image features, in particular we focus on the $\kappa$ power spectrum and the 9 moments described in equation (\ref{par:moments}). Figure \ref{fig:psResiduals} shows the auto power spectra of $\kappa_{\rm born},\kappa_{\rm ll},\kappa_{\rm geo}$ as defined in equation (\ref{sim:pb2}). At lowest order the lensing potential contributes to the $\kappa$ power spectrum quadratically. The first post--Born correction to $P^{\kappa\kappa}$ comes in at $O(\Phi^3)$ and is given by
\begin{equation}
\label{res:powercorr}
P^{\kappa\kappa} = P^{\kappa\kappa}_{\rm born,born} + 2P^{\kappa\kappa}_{\rm born,ll} + 2P^{\kappa\kappa}_{\rm born,geo} + O(\Phi^4),
\end{equation}
%
where we defined the $\kappa$ cross spectra as
\begin{equation}
\label{res:powercross}
\langle\tilde{\kappa}_a(\pmb{\ell})\tilde{\kappa}_b(\pmb{\ell}')\rangle = (2\pi)^2\delta^D(\pmb{\ell}+\pmb{\ell}')P_{a,b}^{\kappa\kappa}(\ell).
\end{equation} 
%
Figure \ref{fig:psResiduals} shows the magnitudes of the $O(\Phi^3)$ post--Born corrections to the convergence power spectrum. A similar analysis can be carried out for the various skewness and kurtosis moments of the $\kappa$ field. The main contributions to the skewness and kurtosis enter at $O(\Phi^3)$ and $O(\Phi^4)$, respectively. The first post--Born corrections, on the other hand, enter at $O(\Phi^5),O(\Phi^6)$. In more detail, we have

@@

\section{Discussion}
\label{sec:discuss}
% 
In this section, we discuss our main findings. Figure \ref{fig:csample} shows that the dominant post--Born correction to the $\kappa$ reconstruction comes from the geodesic term (eq.~\ref{sim:gp}). This can be seen by looking at the lens-lens term, which is over an order of magnitude smaller than the geodesic term, and by looking at the full $\kappa-\kappa_{\rm born}$ residuals, which show a very similar overall level, as well as structural detail, to the ones found in the $\kappa_{\rm geo}$ map. \textbf{\color{red} When considering a real observation, in which $\kappa$ is estimated from source ellipticities, one should consider an additional reduced shear correction to $\kappa$, which is of order $O(\Phi^2)$ \citep{HirataKrause} and is proportional to $\gamma\kappa$. We estimated this correction to be one order of magnitude smaller than the dominant geodesic term}. The reader might have noticed the curious alignment feature in Figure \ref{fig:csample}, in which all the dipolar structures in $\kappa_{\rm geo}$ appear to be horizontally aligned throughout the field of view. We believe that this alignment is not a numerical effect, but is originated physically by the $\nabla\delta$ terms in eq.~(\ref{sim:gp}). Early in the line--of--sight integration, the field of view covers a very small section of the simulation box ($\sim7\,{\rm Mpc}$ for the first lens), within which dipolar structures are aligned. These alignments survive after the full integration is completed due to the $\nabla\delta\cdot\nabla\Phi$ term, which couples lenses at different redshifts. To provide evidence that the closest lenses to the observer are responsible for the alignment features, we performed the line--of--sight integration for $\kappa_{\rm geo}$ by rotating the first 5 lenses by 90 degrees. This is sufficient to coherently flip the dipolar structures from horizontal to vertical throughout the field of view.

@@

\begin{itemize}
\item The lowest--order post--Born corrections to the convergence power spectrum, skewness and kurtosis are sufficient to fully account for the discrepancies between ray--tracing and the Born approximation.
\item Using the \ttt{LensTools} software package, Born integration costs 4 times less than ray--tracing, and consumes about half as much memory.
\item The Born approximation for the $\kappa$ power spectrum leads to negligible parameter bias for an LSST--like survey, even in the noiseless case.
\item Fitting an observation with Born--approximated higher $\kappa$ moments leads to significant bias in the $(\Omega_m,w_0,\sigma_8)$ triplet in the noiseless case. However, this bias becomes negligible when galaxy shape noise is added, due to an increased scatter in parameter estimates. This conclusion holds even for future space--based surveys that can probe galaxy angular densities as high as $60\,{\rm arcmin}^{-2}$.    
\end{itemize}
%
In this work we examined the validity of the Born approximation for WL galaxy surveys, but a similar study could in principle be carried for the lensing potential reconstruction from CMB temperature and polarization data \citep{CMBCalabrese}. Lensing of the CMB probes structures over a wider range of redshifts and hence CMB lensing observables can be expected to be more Gaussian distributed than their galaxy lensing counterparts. This suggests the possibility that non--Gaussian features in CMB lensing data could come from post--Born corrections of $O(\Phi^2)$ rather than inherent non--Gaussianity in the Born $O(\Phi)$ terms. This possibility has been suggested by \citep{CMBPrattenLewis}, who looked at the CMB lensing $\kappa$ bi--spectrum. Because of the high significance with which non--Gaussianities in the CMB lensing potential can be detected with future Stage IV experiments \citep{CMBLNG}, we propose to investigate post--Born corrections to CMB lensing observables in future work.
